A Software Defined Radio Based on the Goertzel Algorithm for the DCF77 Longwave Time Signal

### 2024-08-30 paper/main.tex
First round of fixing typos completed

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
%Template version as of 6/27/2024

@@

\begin{document}
\title{Decoding the Amplitude-Modulated Part of the DCF77 Longwave Time Signal by using the Goertzel Algorithm}

\author{\IEEEauthorblockN{1\textsuperscript{st} Leander Hackmann}
\IEEEauthorblockA{\textit{Master Embedded Systems Engineering} \\
\textit{FH Dortmund}\\
Dortmund, Germany \\
Mat.-Nr.: 7217912, leander.hackmann001@stud.fh-dortmund.de}
}

@@

\section{Introduction}
Radio controlled clocks are very popular because they dont require periodical re-adjustment in comparison to common mechanical or quartz-powered clocks.
Instead, the information about the current time is obtained from a RF-signal, transmitted by an official authority that ensures its availability and correctness.
The probably most important time signal in Europe is transmitted by the station DCF77 from Mainflingen, a city in Germany that is close to Frankfurt am Main.
The station is operated by a private company that is commissioned by the Physikalisch-Technische Bundesanstalt (PTB) which also supplies the current time information,
generated by two caesium clocks and two caesium fountains.
The time signal's operating frequency lies in the long wave area at \SI{77.5}{\kilo\hertz}, simple modulation techniques (amplitude and phase-modulation) are used and it is transmitted with \SI{100}{\kilo\watt} Effective Isotropic Radiated Power (EIRP) to enable the reception of the
signal in relatively close distances (up to \SI{2000}{\kilo\meter}) with simple technical equipment \cite{b2}.
\begin{figure}[!htbp]
    \centerline{\includegraphics[width=0.3\textwidth]{img/dcf77_range.jpg}}
    \caption{Coverage of the DCF77 signal. Adapted from \cite{b1}}
    \label{fig:dcf77_range}
\end{figure}
\FloatBarrier
End-user devices usually make use of the amplitude-modulated part of the signal by using an amplitude modulation receiver, often a by employing specialized integrated circuits,
together with a rather big ferrite-rod antenna to obtain the time code signal.
This time code signal then needs further hardware, often in form of a microcontroller used as the applicaton processor, to decode and retrieve the actual time information.
With the increasing availability of computation power and Digital Signal Processing (DSP) capabilities inside cheap mixed-signal microcontrollers, which are already used
to process the time information, the idea to replace to replace the analog receiver portion with a Software Defined Radio (SDR) is more or less obvious.
Moving the receiver into the application processor would allow for a simpler hardware design by shrinking down the complexity and the parts count of the
external analog circuitry. In fact, only a tuned antenna with a basic pre-amplifier is needed.
\par
In this paper, a straightfoward principle for designing a lightweight receiver for the DCF77, based on SDR-techniques, is presented.
After taking a look at the structure of the DCF77 signal, the theoretical operation principle of the receiver is explained and shown in a MATLAB simulation.
To further underline the simplicity and the possibility to implement the principle on a system with limited resources, a demonstration on a Arm Cortex-M3 microcontroller
is shown.

@@

\section{Principle of the receiver}
As the seen in the description of the DCF77 signal, retrieving back the binary time signal from the amplitude-modulated part is more or less trivial.
Conventional applications execute this task by using analog receiving circuits, usually by using a simple amplitude modulation receiver together with a
tuned ferrite-rod antenna, acting as a Bandpass Filter (BPF), to filter out the irrelevant part of the spectrum. 
Moving the receiver into the digital domain allows for more advanced techniques to retrieve this binary time signal, rather than demodulating the signal with a
diode detector inside the amplitude modulation receiver.
Because the information is contained only in the amplitude of the signal, and therefore inside the signal strength at the receiver, it can also be obtained by doing a
continuous analysis of the spectrum at the carrier's frequency of \SI{77.5}{\kilo\hertz}.
This task will be carried out by using the Goertzel algorithm, which is presented later.
The structure of the whole receiver then simplifies to only two external hardware parts
(besides the application processor) and four internal software parts.

@@

\par
Just as with the conventional FFT, the amount of needed calculations for executing the Goertzel algorithm is based on the number of samples.
In total, $N + 1$ multiplications and $2 \cdot N + 2$ additions are needed, which can be seen from the formulas for the IIR-part and the final calculation of the squared magnitude.
For example, a length of processing $N = 512$ samples requires in $513$ multiplications and $1026$ additions to calculate the squared magnitude of the desired frequency.
In comparison, the typical radix-2 FFT, needs $\frac{N}{2} \cdot \log_{2}(N)$ complex multiplications and $N \cdot \log_{2}(N)$ complex additions to be completed. With $N = 512$ this would result in $2034$ complex multiplications and $4608$ complex additions.
\par
With this simple example, it is already visible that the Goertzel algorithm can reduce the amount of needed resources compared to a full FFT, when only one or a small amount of frequency bins should be analyzed.
It is therefore used to observe the DCF77 signal's amplitude over time by being executed periodically.
As the gap of the encoded symbols is eiter \SI{100}{\milli\second} or \SI{200}{\milli\second} long, the duration of the analyzation set to \SI{10}{\milli\second} to provide a good tradeoff between resolution in time and a longest possible analyzation time
to leverage the DFT gain.
The square roots of the results are taken to finally calculate the magnitude of the DCF77 signal's amplitude and then passed to a detector to obtain the binary time signal.
Because of the duration of the analyzation being equal to \SI{10}{\milli\second}, the sampling rate after the Goertzel algorithm is equal to:
\begin{equation}
    f_{s,g} = \frac{1}{\SI{10}{\milli\second}} = \SI{100}{\hertz}
\end{equation}

@@

\subsection{Results}
Multiple simulations with different values for the SNR starting from \SIrange{5}{-7}{\deci\bel} were executed to verify and evaluate the receiver's capabilities to reconstruct the DCF77's time code signal in different receiving conditions.
From these simulations, the most relevant results occured in the case where the SNR was equal to \SI{-5}{\deci\bel}, because this was the lowest SNR where the reciver was still able to
retrieve back the time code signal without errors.
The frequency domain plots of the two FFTs, shown in Fig. \ref{fig:sim_fd_m5dB}, verify the DCF77 signal's aliases to show up exactly at \SI{\pm 5.5}{\kilo\hertz}.
From the plot showing the DCF77 signal after passing the AWGN-channel, it can be seen that the noise floor is located at approximately \SI{-50}{\deci\bel} while the peaks
of the DCF77 signaö have a value of approximately \SI{-8}{\deci\bel}.
The difference of \SI{42}{\deci\bel} should not be confused with the SNR, as the power of the noise added by the AWGN-Channel is spread evenly over all frequencies.
The correct SNR was verified to be exactly \SI{-5}{\deci\bel} by using MATLAB's built-in command "snr()" to calculate the SNR by comparing the before and after adding noise.
\begin{figure}[!htbp]
    \centerline{\includegraphics[width=0.4\textwidth]{img/sim_fd_m5dB.eps}}
    \caption{Plots of the frequency domain at a SNR of \SI{-5}{\deci\bel}}
    \label{fig:sim_fd_m5dB}
\end{figure}
\FloatBarrier
The time domain plots of Fig. \ref{fig:sim_td_m5dB} show the creation of the DCF77 signal and the internal operation of the receiver.
The first plot shows the time code signal that should be retrieved back by the receiver to allow a visual verification that the DCF77 signal is modulated retrieved back correctly.
It can be seen that each symbol has a length of \SI{1}{\second} with a gap of either \SI{200}{\milli\second} or \SI{100}{\milli\second} at the beginning.
From the second plot, the amplitude can be seen to switch between $0.15$ and $1$, representing the changes between \SI{15}{\percent} and \SI{100}{\percent}.
The third plot shows the DCF77 signal with added noise and an average amplitude of approximately $3.6$ with the orignal waveform not being visible anymore.
The fourth plot shows the results of the Goertzel algorithm segments together wit the value of the calculated threshold.
The operation of the Goertzel algorithm can be verified, as the calculated magnitude of the frequency bin is clearly following the waveform of the time code signal.
The value propagation of the threshold is as slow as expected with the value only experiencing minor drops during the gaps of the signal.
A value where the detector starts to take valid decisions is reached after approximately \SI{5}{\second}.
The threshold then approaches asymptotically the value where it is exactly located in the middle between the area of values representing an apmlitude of \SI{15}{\percent} and
the area of values representing \SI{100}{\percent}.
As the value of the EMA is proportional to the threshold, the plot is left out.
In the fifth plot, the retrieved time code signal is visualized.
It is obvious that the signal contains errors until the threshold has reached a level where valid decisions are taken by the detector.
After \SI{5}{\second}, when the threshold reached this value, no more errors occured.
\par
All simulation runs with a smaller SNR contained an increasing amount of errors, caused by the noise to raise or drop single points above or below the threshold, placing them in the
area where they were detected to correspond to the opposite symbol.
This effect can be imagined by thinking of the height of the area where the results are distributed by the noise is increased until both areas touch the threshold and start to overlap.
Single points ending up in the other area are then interpreted as the wrong symbol.

\begin{figure*}[!htbp]
    \centerline{\includegraphics[width=1.1\textwidth]{img/sim_td_m5dB.eps}}
    \caption{Plots of the time domain at a SNR of \SI{-5}{\deci\bel}}
    \label{fig:sim_td_m5dB}
\end{figure*}

\section{Demo implementation}
The receiver's structure, shown in Fig. \ref{fig:block_diag_receiver}, was implemented on a STM32F103C8T6, an Arm Cortex-M3 microcontroller, to provide a brief proof of concept.
To for an optical verification, the obtained time code signal is used to turn on and off a Light-Emitting Diode (LED), but this signal
could also be used to decode the DCF77's transmitted time data.
As a tuned antenna together with a pre-amplifier is needed to filter out the DCF77 signal and amplify it to a level where it can be measured by the microcontroller's internal ADC, a minimalistic circuit 
consisting of a tank circuit tuned to \SI{77.5}{\kilo\hertz} and a two-stage transistor amplifier was built based on the circuit shown in \cite{b12}.
The rest of the of the receiver's structure was implemented in software using C and single-precision floating-point for the caluclations.
\par
A timer was set up to create an internal Pulse Width Modulation (PWM) signal to trigger the conversion of the microcontroller's internal ADC at the sampling frequency of \SI{24}{\kilo\hertz}.
The samples are transferred using circular Direct Memory Access (DMA) into a buffer of the size of $480$.
This size was chosen because it is equal to twice the amount of samples that are needed for the analyzation of one segment.
During runtime, the DMA hardware module fills this buffer and two interrupts are executed when the buffer was filled half way and when it is completely full.
Because of the circular setup of the DMA, it fills the buffer continuously and wraps around back to the first address, after reaching the buffer's end.
This allows a continuous operation of the application with a precise sampling frequency while eliminating the need to handle ADC conversions manually. 
As a segment is ready to be analyzed with every interrupt, the operations needed to process the signal processing are executed inside this Interrupt Service Routine (ISR).
The result of each segment after the detector is then used to turn on or off a General Purpose Input Output (GPIO) that is connected to a LED.
\par
Although its rather simplistic implementation, the demo implementation worked right away.
The behavior of the receiver that was observed in the simulation, first outputting a series of wrong results when adapting to the
current receiving conditions, is also noticable when running the demo application.
After powerup or changes in the environment and therefore during the stabilization of the threshold to its new value, the LED shows
arrythmic behavior for approximately \SI{5}{\second} before the application continues to operate correctly.
Other than that, the demo implementation shows that the presented principle also works in reality and that it can be implemented
with low effort.
Based on this structure, the code can be expanded to a full clock by decoding the information from the retrieved time code signal.
The code is available in the paper's GitHub repository \cite{b13}.

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
%Template version as of 6/27/2024

@@

\begin{document}
\title{A Software Defined Radio Based on the Goertzel Algorithm for the DCF77 Longwave Time Signal}

\author{\IEEEauthorblockN{1\textsuperscript{st} Leander Hackmann}
\IEEEauthorblockA{
\textit{Master Embedded Systems Engineering}\\
\textit{FH Dortmund}\\
Dortmund, Germany\\
Mat.-Nr.: 7217912, leander.hackmann001@stud.fh-dortmund.de\\}
}

@@

\section{Introduction}
Radio controlled clocks are very popular because they do not require periodical re-adjustment in comparison to common mechanical or quartz-powered clocks.
Instead, the information about the current time is obtained from a RF-signal, transmitted by an official authority that ensures its availability and correctness.
The probably most important time signal in Europe is transmitted by the station DCF77 from Mainflingen, a city in Germany that is close to Frankfurt am Main.
The station is operated by a private company that is commissioned by the Physikalisch-Technische Bundesanstalt (PTB), which also supplies the current time information,
generated by two caesium clocks and two caesium fountains.
The time signal's operating frequency lies in the long wave area at \SI{77.5}{\kilo\hertz}, simple modulation techniques (amplitude and phase-modulation) are used and it is transmitted with \SI{100}{\kilo\watt} Effective Isotropic Radiated Power (EIRP) to enable the reception of the
signal in relatively close distances (up to \SI{2000}{\kilo\meter}) with simple technical equipment \cite{b2}.
\begin{figure}[!htbp]
    \centerline{\includegraphics[width=0.3\textwidth]{img/dcf77_range.jpg}}
    \caption{Coverage of the DCF77 signal. Adapted from \cite{b1}}
    \label{fig:dcf77_range}
\end{figure}
\FloatBarrier
End-user devices usually make use of the amplitude-modulated part of the signal by using an amplitude modulation receiver, often a by employing specialized integrated circuits,
together with a rather big ferrite-rod antenna to obtain the time code signal.
This time code signal then needs further hardware, often in form of a microcontroller used as the applicaton processor, to decode and retrieve the actual time information.
With the increasing availability of computation power and Digital Signal Processing (DSP) capabilities inside cheap mixed-signal microcontrollers, which are already used
to process the time information, the idea to replace the analog receiver portion with a Software Defined Radio (SDR) is more or less obvious.
Moving the receiver into the application processor would allow for a simpler hardware design by shrinking down the complexity and the parts count of the
external analog circuitry. In fact, only a tuned antenna with a basic pre-amplifier is needed.
\par
In this paper, a straightfoward principle for designing a lightweight receiver for the DCF77, based on SDR-techniques, is presented.
After taking a look at the structure of the DCF77 signal, the theoretical operation principle of the receiver is explained and shown in a MATLAB simulation.
To further underline the simplicity and the possibility to implement the principle on a system with limited resources, a demonstration on a Arm Cortex-M3 microcontroller
is shown.

@@

\section{Principle of the receiver}
As seen in the description of the DCF77 signal, retrieving back the binary time signal from the amplitude-modulated part is more or less trivial.
Conventional applications execute this task by using analog receiving circuits, usually by using a simple amplitude modulation receiver together with a
tuned ferrite-rod antenna, acting as a Bandpass Filter (BPF), to filter out the irrelevant part of the spectrum. 
Moving the receiver into the digital domain allows more advanced techniques to retrieve this binary time signal, rather than demodulating the signal with a
diode detector inside the amplitude modulation receiver.
Because the information is contained only in the amplitude of the signal, and therefore inside the signal strength at the receiver, it can also be obtained by doing a
continuous analysis of the spectrum at the carrier's frequency of \SI{77.5}{\kilo\hertz}.
This task will be carried out by using the Goertzel algorithm, which is presented later.
The structure of the whole receiver then simplifies to only two external hardware parts
(besides the application processor) and four internal software parts.

@@

\par
Just as with the conventional FFT, the amount of needed calculations for executing the Goertzel algorithm is based on the number of samples.
In total, $N + 1$ multiplications and $2 \cdot N + 2$ additions are needed, which can be seen from the formulas for the IIR-part and the final calculation of the squared magnitude.
For example, a length of processing $N = 512$ samples requires in $513$ multiplications and $1026$ additions to calculate the squared magnitude of the desired frequency.
In comparison, the typical radix-2 FFT, needs $\frac{N}{2} \cdot \log_{2}(N)$ complex multiplications and $N \cdot \log_{2}(N)$ complex additions to be completed. With $N = 512$ this would result in $2034$ complex multiplications and $4608$ complex additions.
\par
With this simple example, it is already visible that the Goertzel algorithm can reduce the amount of needed resources compared to a full FFT, when only one or a small amount of frequency bins should be analyzed.
It is therefore used to observe the DCF77 signal's amplitude over time by being executed periodically.
As the gap of the encoded symbols is eiter \SI{100}{\milli\second} or \SI{200}{\milli\second} long, the duration of the analyzation is set to \SI{10}{\milli\second} to provide a good tradeoff between resolution in time and a longest possible analyzation time
to leverage the DFT gain.
The square roots of the results are taken to finally calculate the magnitude of the DCF77 signal's amplitude and then passed to a detector to obtain the binary time signal.
Because of the duration of the analyzation being equal to \SI{10}{\milli\second}, the sampling rate after the Goertzel algorithm is equal to:
\begin{equation}
    f_{s,g} = \frac{1}{\SI{10}{\milli\second}} = \SI{100}{\hertz}
\end{equation}

@@

\subsection{Results}
Multiple simulations with different values for the SNR starting from \SIrange{5}{-7}{\deci\bel} were executed to verify and evaluate the receiver's capabilities to reconstruct the DCF77's time code signal in different receiving conditions.
From these simulations, the most relevant results occured in the case where the SNR was equal to \SI{-5}{\deci\bel}, because this was the lowest SNR where the reciver was still able to
retrieve back the time code signal without errors.
The frequency domain plots of the two FFTs, shown in Fig. \ref{fig:sim_fd_m5dB}, verify the DCF77 signal's aliases to show up exactly at \SI{\pm 5.5}{\kilo\hertz}.
From the plot showing the DCF77 signal after passing the AWGN-channel, it can be seen that the noise floor is located at approximately \SI{-50}{\deci\bel} while the peaks
of the DCF77 signal have a value of approximately \SI{-8}{\deci\bel}.
The difference of \SI{42}{\deci\bel} should not be confused with the SNR, as the power of the noise added by the AWGN-Channel is spread evenly over all frequencies.
The correct SNR was verified to be exactly \SI{-5}{\deci\bel} by using MATLAB's built-in command "snr()" to calculate the SNR by comparing the before and after adding noise.
\begin{figure}[!htbp]
    \centerline{\includegraphics[width=0.4\textwidth]{img/sim_fd_m5dB.eps}}
    \caption{Frequency domain plots of the simulation run with a SNR of \SI{-5}{\deci\bel}}
    \label{fig:sim_fd_m5dB}
\end{figure}
\FloatBarrier
The time domain plots of Fig. \ref{fig:sim_td_m5dB} show the creation of the DCF77 signal and the internal operation of the receiver.
The first plot shows the time code signal that should be retrieved back by the receiver to allow a visual verification that the DCF77 signal is modulated retrieved back correctly.
It is visible that each symbol has the expected length of \SI{1}{\second} with a gap of either \SI{200}{\milli\second} or \SI{100}{\milli\second} at the beginning.
From the second plot, the amplitude can be seen switching between $0.15$ and $1$, representing the changes between \SI{15}{\percent} and \SI{100}{\percent}.
The third plot shows the DCF77 signal with added noise and an average amplitude of approximately $3.6$ with the orignal waveform not being visible anymore.
The fourth plot shows the results of the Goertzel algorithm segments together wit the value of the calculated threshold.
The operation of the Goertzel algorithm can be verified, as the calculated magnitude of the frequency bin is clearly following the waveform of the time code signal.
The value propagation of the threshold is as slow as expected with the value only experiencing minor drops during the gaps of the signal.
A value where the detector starts to take valid decisions is reached after approximately \SI{5}{\second}.
The threshold then approaches asymptotically the value where it is exactly located in the middle between the area of values representing an apmlitude of \SI{15}{\percent} and
the area of values representing \SI{100}{\percent}.
As the value of the EMA is proportional to the threshold, the plot is left out.
In the fifth plot, the retrieved time code signal is visualized.
It is obvious that the signal contains errors until the threshold has reached a level where valid decisions are taken by the detector.
After \SI{5}{\second}, when the threshold reached this value, no more errors occured.
\par
All simulation runs with a smaller SNR contained an increasing amount of errors, caused by the noise to raise or drop single points above or below the threshold, placing them in the
area where they were detected to correspond to the opposite symbol.
This effect can be imagined by thinking of the height of the area where the results are distributed by the noise is increased until both areas touch the threshold and start to overlap.
Single points ending up in the other area are then interpreted as the wrong symbol.

\begin{figure*}[!htbp]
    \centerline{\includegraphics[width=1.1\textwidth]{img/sim_td_m5dB.eps}}
    \caption{Time domain plots of the simulation run with a SNR of \SI{-5}{\deci\bel}}
    \label{fig:sim_td_m5dB}
\end{figure*}

\section{Demo implementation}
The receiver's structure, shown in Fig. \ref{fig:block_diag_receiver}, was implemented on a STM32F103C8T6, an Arm Cortex-M3 microcontroller, to provide a brief proof of concept.
For an optical verification, the obtained time code signal is used to turn on and off a Light-Emitting Diode (LED), but this signal
could also be used to decode the DCF77's transmitted time data.
As a tuned antenna together with a pre-amplifier is needed to filter out the DCF77 signal and amplify it to a level where it can be measured by the microcontroller's internal ADC, a minimalistic circuit 
consisting of a tank circuit tuned to \SI{77.5}{\kilo\hertz} and a two-stage transistor amplifier was built based on the circuit shown in \cite{b12}.
The rest of the of the receiver's structure was implemented in software using C and single-precision floating-point for the caluclations.
\par
A timer was set up to create an internal Pulse Width Modulation (PWM) signal to trigger the conversion of the microcontroller's internal ADC at the sampling frequency of \SI{24}{\kilo\hertz}.
The samples are transferred using circular Direct Memory Access (DMA) into a buffer of the size of $480$.
This size was chosen because it is equal to twice the amount of samples that are needed for the analyzation of one segment.
During runtime, the DMA hardware module fills this buffer and two interrupts are executed when the buffer was filled half way and when it is completely full.
Because of the circular setup of the DMA, it fills the buffer continuously and wraps around back to the first address, after reaching the buffer's end.
This allows a continuous operation of the application with a precise sampling frequency while eliminating the need to handle ADC conversions manually. 
As a segment is ready to be analyzed with every interrupt, the operations needed to process the signal processing are executed inside this Interrupt Service Routine (ISR).
The result of each segment after the detector is then used to turn on or off a General Purpose Input Output (GPIO) that is connected to a LED.
\par
Although its rather simplistic implementation, the demo implementation worked right away.
The behavior of the receiver that was observed in the simulation, first outputting a series of wrong results when adapting to the
current receiving conditions, is also noticable when running the demo application.
After powerup or changes in the environment and therefore during the stabilization of the threshold to its new value, the LED shows
arrythmic behavior for approximately \SI{5}{\second} before the application continues to operate correctly.
Other than that, the demo implementation shows that the presented principle also works in reality and that it can be implemented
with low effort.
Based on this structure, the code can be expanded to a full clock by decoding the information from the retrieved time code signal.
The code is available in the paper's GitHub repository \cite{b13}.
